Continuation Passing Style for Effect Handlers

### 2020-01-07 docs/solution.tex
Define syntax and operational semantics in article

\documentclass{article}
\usepackage[utf8]{inputenc}
% \usepackage{polski}
\usepackage[]{amsthm}
\usepackage[]{amssymb} %gives us the character \varnothing

@@

    We provide an implementation of algebraic effects and handlers by doing continuation
    passing style tranformation of the functional programming language Freak,
    which is based on the existing Links langauge \cite{handlers-cps}.

@@

In this article we present experimental programming language Freak, which is an
implementation of Continuation Passing Style for Effect Handlers paper \cite{handlers-cps},
with addition of a few basic constructs. We start by presenting the related work,
then discuss syntax and operational semantics. Basic usage guide for playing with
the language is provided, as well as implementation details and examples. We conclude
by stating what are the possible augmentations, that are intended to be made in the future.

@@

@@

    The source language's dynamics has been described
    extensively by providing small-step operational semantics,
    continuation passing style transformation \cite{handlers-cps} as well
    as abstract machine \cite{liberating-effects}, which was proved to coincide
    with CPS translation. That being said, Freak introduces new basic
    constructs to the language, for which we shall define the semantics.

    \subsection{Integers}

    \subsection{Binary operators}

@@

    The Freak implementation is available at \href{https://github.com/Tomatosoup97/freak}{https://github.com/Tomatosoup97/freak}.
    While the paper provided a good overview on the language and the translation,
    the lower-level details were ommited. Therefore, the actual CPS implementation
    is based on A. Appel \cite{appel-continuations}, which provides a translation
    for a simplified ML calculus. \\

    The core data structure into which the source program is transformed,
    the \verb!Continuation Computation!, is defined as follows:

@@

    Type system as of this day is not implemented, as the focus has been put
    on CPS transformation. Further work is required here, especially considering
    the fact that a huge advantage of algebraic effects is that they are explicitly
    defined in the type of a computation.

@@

\section{Examples}

    In this section we present a few examples to show the capabilities of the
    language. The ideas has been based on \cite{programming-in-eff}, and thus will not be
    described in great details. More exemplary programs in Freak language can
    be found \href{https://github.com/Tomatosoup97/freak/tree/master/src/programs}{\underline{here}}.

@@

    The Freak language is experimental and a lot of possible enhancements
    could be adopted. Here we provide a list of proposed directions in which
    further work could be made.

@@

    \subsection{Make the langugae more usable}

    While the language is turing-complete, for convenient usage it
    requires more basic constructs and syntactic sugar for common patterns
    that would ease the programming.

@@

    The Freak language is limited to a single instance of an effect. We would
    need to support cases where many instances of the algebraic effects, with
    the same handlers code, could be instantiated. The current state of the
    art introduces a concept of resources and instances, as in Eff \cite{programming-in-eff},
    or instance variables, as in Helium \cite{binders-labels}.

@@

    Other languages, like Koka \cite{leijen-koka}, or even core of the Links, are
    performing selective CPS translation, which reduces the overhead on code
    that does not perform algebraic effects. Our current translation is fully
    embeded in the CPS.

@@

    Shallow and deep handlers while being able to simluate each other up to
    administrative reductions, have a very different meaning from a theoretical
    point of view. Implementing them as defined by Lindley et al. \cite{shallow-handlers} could
    be another way of enhancing Freak.

\section{Conclusions}

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
% \usepackage{polski}
\usepackage[]{amsthm}
\usepackage{amsfonts}
\usepackage[]{amssymb}
\usepackage{amsmath}
\usepackage{syntax}
\usepackage{mathalfa}

@@

    We provide an implementation of algebraic effects and handlers by doing continuation
    passing style transformation of the functional programming language Freak,
    which is based on the existing Links language \cite{handlers-cps}.

@@

In this article we present experimental programming language Freak, which is an
implementation of Continuation Passing Style for Effect Handlers paper \cite{handlers-cps},
with the addition of a few basic constructs. We start by presenting the related work,
then discuss syntax and operational semantics. Basic usage guide for playing with
the language is provided, as well as implementation details and examples. We conclude
by stating what are the possible augmentations, that are intended to be made in the future.

@@

    The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
    $V \oplus W$ and $V \approx W$ are respectively binary and relational
    operators, where we support basic arithmetic and comparison operations.
    \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ is a standard branching statement.
    The other constructs are just as in Links, with slight syntax modifications.
    Actual programs in Freak can be found in \fullref{sec:examples}

    \begin{grammar}

        <Values V, W> ::= $ x $ | $nat \; n$ \\
            | $ \backslash x : A \rightarrow M $ | \textbf{rec} $ g \; x \rightarrow M $\\
            | $V \oplus W$ | $V \approx W$ \\
            | <> | $ \{ \ell = V; W\} $  | $[ \ell \; V]^{R}$

        <Computations M, N> ::= $ V $ $ W $ \\
            | \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ \\
            | \textbf{let} $\{\ell  = x; y\} = V$ \textbf{in} $ N $ \\
            | \textbf{case} $V \{ \ell \; x \rightarrow M; y \rightarrow N\}$ | \textbf{absurd} $ V $ \\
            | \textbf{return} $V$ | \textbf{let} $ x \leftarrow M $ \textbf{in} $ N $ \\
            | \textbf{do} $\ell \; V$ | \textbf{handle} $M$ \textbf{with} $ \{ H \} $

        <Handlers H> ::= \textbf{return} $ x \rightarrow M $ | $ \ell \; p \; r \rightarrow M, H $

        <Binary operators $\oplus$> ::= + | - | * | /

        <Relational operators $\approx$> ::= $ \textless $ | $\leqslant$ | $>$ | $\geqslant$ | == | !=

    \end{grammar}

@@

    The source language's dynamics have been described
    extensively by providing small-step operational semantics,
    continuation passing style transformation \cite{handlers-cps} as well
    as abstract machine \cite{liberating-effects}, which was proved to coincide
    with CPS translation. That being said, Freak introduces new basic
    constructs to the language, for which we shall define the semantics.

    \begin{flushleft}
    Extension of the evaluation contexts:\\
    \end{flushleft}

    \begin{flushleft}
    $\mathcal{E}$ ::= $ \mathcal{E} \oplus W \; | \; V \oplus \mathcal{E} \; |$ \textbf{if} $\mathcal{E}$ \textbf{then} $M$ \textbf{else} $N$ \\
    \end{flushleft}

    \begin{flushleft}
    Small-step operational semantics:
    \end{flushleft}

    \begin{flushleft}
    \textbf{if} $nat \; n$ \textbf{then} $M$ \textbf{else} $N \rightsquigarrow M \quad \quad $  if $n \neq 0$ \\
    \textbf{if} $nat \; n$ \textbf{then} $M$ \textbf{else} $N \rightsquigarrow N \quad \quad $    if $n = 0$ \\
    \end{flushleft}

    \begin{flushleft}
    $nat \; n \oplus nat \; n' \rightsquigarrow n'' \quad \quad $    if $ n'' = n \oplus n' $ \\
    $nat \; n \approx nat \; n' \rightsquigarrow 1  \quad \quad $    if $ n \approx n' $ \\
    $nat \; n \approx nat \; n' \rightsquigarrow 0  \quad \quad $    if $ n \not\approx n' $

    \end{flushleft}

@@

    The Freak implementation is available at \href{https://github.com/Tomatosoup97/freak}{https://github.com/Tomatosoup97/freak}.
    While the paper provided a good overview of the language and the translation,
    the lower-level details were omitted. Therefore, the actual CPS implementation
    is based on A. Appel \cite{appel-continuations}, which provides a translation
    for a simplified ML calculus. \\

    Abstract Syntax Tree structure is defined just as syntax is. The core data
    structure into which the source program is transformed,
    the \verb!Continuation Computation!, is defined as follows:

@@

    The type system as of this day is not implemented, as the focus has been put
    on CPS transformation. Further work is required here, especially considering
    the fact that a huge advantage of algebraic effects is that they are explicitly
    defined in the type of a computation.

@@

\section{Examples}
\label{sec:examples}

    In this section we present a few examples to show the capabilities of the
    language. The ideas have been based on \cite{programming-in-eff}, and thus will not be
    described in great details. More exemplary programs in Freak language can
    be found \href{https://github.com/Tomatosoup97/freak/tree/master/src/programs}{\underline{here}}.

@@

    The Freak language is experimental and a lot of possible enhancements
    could be adopted. Here we provide a list of the proposed directions in which
    further work could be done.

@@

    \subsection{Make the language more usable}

    While the language is Turing-complete, for convenient usage it
    requires more basic constructs and syntactic sugar for common patterns
    that would ease the programming.

@@

    The Freak language is limited to a single instance of an effect. We would
    need to support cases where many instances of the algebraic effects, with
    the same handler code, could be instantiated. The current state of the
    art introduces a concept of resources and instances, as in Eff \cite{programming-in-eff},
    or instance variables, as in Helium \cite{binders-labels}.

@@

    Other languages, like Koka \cite{leijen-koka}, or even the core of the Links, are
    performing selective CPS translation, which reduces the overhead on code
    that does not perform algebraic effects. Our current translation is fully
    embedded in the CPS.

@@

    Shallow and deep handlers while being able to simulate each other up to
    administrative reductions, have a very different meaning from a theoretical
    point of view. Implementing them as defined by Lindley et al. \cite{shallow-handlers} could
    be another way of enhancing Freak.
